Linting Linearity in Core/System FC

### 2022-11-22 proposal.tex
Proposal introduction and foreword

\usepackage{cmll}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

@@

@@

\subsection{Motivation}

\begin{itemize}
    \item Are we really preserving linearity?
    \item Inform/unlock other optimisations that take into account linearity
\end{itemize}

\section{Background}

\begin{itemize}
    \item Linear types
    \item Linear Haskell
    \item Core
    \item Sistemas de inferência
    \item GADTs e Coercions
\end{itemize}

\section{Related Work}

\section{Technical Details}

Since Linear Haskell's\ref{}
publication and implementation release in GHC 9.2, Haskell's type system
supports linearity annotations in functions -- bringing linear types into a
mainstream pure and lazy functional language.

System FC is the formal system in which the implementation of GHC's intermediate
representation language \emph{Core} is based on.

There are at least two distinct typecheckers in core. The first is run on the frontend
language, i.e. the Haskell we write, and is a big and complex typechecker. The
second is run on the intermediate language \emph{Core} that we obtain from
desugaring Haskell.

\emph{Core} is a much smaller and more principled language than the whole of
Haskell (even though we can compile the whole of Haskell to it), and the
typechecker for it is small and fast due to \emph{Core} being explicitly typed
and having a very small abstract syntax tree. This typechecker is called
\emph{Lint} and gives us guarantees of correctness (i.e. sanity) in face of
the complexity of all the transformations a Haskell program undergoes, such as
type inference, desugaring and optimising transformations (and other Core
related passes?).

\begin{itemize}
    \item In GHC, the type Type is shared between the surface language and Core.
        So modifying the types implies that Core is modified as well.

    \item More importantly, because Core is a good check that our implementation
        works as intended. That is, a linearly typed Core will ensure that
        linearly-typed programs are indeed desugared to linearly-typed programs
        in Core. And that optimisations do not destroy linearity.
\end{itemize}

The addition of linear types comes at two levels: frontend and Core.
In the frontend, we can now annotate with linearity type declarations and
typecheck programs that use them accordingly (a linear fuction will consume its
argument exactly once if it is consumed exactly once, for some definition of
\emph{consume} that I should revise here). In Core, we still have linearity
annotations and ideally \emph{Lint} would check that all the GHC transformations
to our linear program preserved its linearity. However, \textbf{it can't}!

Despite the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far from
trivial. Diverse problems with linearity spring up when we get past the
desugarer. In particular, optimizing transformations, coercions from GADTs and
type families, recursive lets, and empty case expressions don't currently fit in
with linearity.

We believe that GHC's transformations are correct, and it is the linear type
system that can't accommodate the resulting programs. We aim to formalise a type
system that is able to type check \emph{Core}/\emph{System FC} at all points in
the core pipeline and implement it into GHC to be able to preserve linearity
accross the stages and to enable \emph{Lint} to preserve our sanity regarding
linearity.

@@

Now, with the following example, my interpretation of the current rule says that
the following case expression consumes the scrutinee twice because the case
binder is used. However, we know for sure that in the branch where z is used, z
is equal to False, and using False doesn't violate linearity since we can
unrestrictedly create nullary data constructors. A situation similar to this one
below happens after the CSE transformation.
\begin{code}
    case <complex expression> $\leadsto U$ of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }

@@

Taking this example (copy example over) we can see that we don't know how to say
that x is indeed linear in one case and unrestricted in the other, even though
it is according to its type. We'd need some sort of coercion to coerce through
the multiplicity to the new one we uncover when we pattern match on the GADT
evidence (...)

>>>>

\usepackage{todonotes}
\usepackage{cmll}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

@@

% \section{Introduction}

% \subsection{Motivation}

% \begin{itemize}
%     \item Are we really preserving linearity?
%     \item Inform/unlock other optimisations that take into account linearity
% \end{itemize}

% \section{Background}

% \begin{itemize}
%     \item Linear types
%     \item Linear Haskell
%     \item Core
%     \item Sistemas de inferência
%     \item GADTs e Coercions
% \end{itemize}

% \section{Related Work}

% \section{Technical Details}

\section{Foreword}

This document is a work in progress proposal that I will deliver at the end of
the semester as part of my master thesis preparation. The document needs a
proper introduction and lengthy background and related work section besides the
content (to be extended) that follows, such that an outsider might understand
what I propose to do. However, this initial iteration is instead targeted at
those already familiar with the problem and serves as a (less bureaucratic)
research proposal on linting linearity in Core and to showcase the progress I
have made so far.

@@

Since Linear Haskell's\cite{} publication and implementation release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types into a mainstream pure and lazy functional language. In concrete,
function arrows can be annotated with a multiplicity. If a function is linear it
must consume its argument exactly once if it is consumed exactly once, for some
definition of \emph{consume} that I should revise here.

Linearly typed programs are proven/checked to be linear by the type system, so
the addition of linear types evidently implied changing the type checker to
support linearity. There exist, however, two distinct type checkers in Haskell.
The first is run on the surface language, i.e. the Haskell we write, and is a
big and complex type checker that now also supports typing linearity. The second
is run on the intermediate language \emph{Core} that we obtain from desugaring
Haskell.

\emph{Core} is a much smaller language than the whole of Haskell (even though we
can compile the whole of Haskell to it!) and the type checker for it is small
and fast due to \emph{Core} being explicitly typed and having a small abstract syntax
tree. This type checker (called \emph{Lint}) gives us guarantees of correctness
(i.e. sanity) in face of all the complex transformations a Haskell program
undergoes such as desugaring and optimising transformations because the linter
is always run on the resulting code before being compiled to untyped machine
code.

% System FC is the formal system in which the implementation of GHC's intermediate
% representation language \emph{Core} is based on.

We want Core to give us guarantees about our desugaring and transformations with
regard to linearity too -- a linearly typed Core will ensure that linearly-typed
programs remain correct (i.e.  linearity is preserved) after desugaring and all
GHC transformations (optimisations shouldn't destroy linearity).

Core is already annotated with linearity, but \textbf{we currently ignore it}!
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far from
trivial, and the implemented formal system cannot accomodate various
optimising transformations and from them resulting programs, that seemingly
violate linearity, but which do preserve it semantically. Therefore, we reject
various valid programs with regard to linearity after desugaring: the current
solution, because disabling optimisations can inccur great performance costs, is
to disable the linear linter. We believe that GHC's transformations are correct,
and it is the linear type system that can't accommodate the resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and the
% linter rejects valid programs which is quite undesirable.

% For example, optimizing transformations, coercions and type families, recursive
% lets, and empty case expressions don't currently fit in with linearity.

We propose to formalise a type system that is able to type check linear
\emph{Core} at all points in the core pipeline and then to implement it into
GHC. Proposal... \todo{proposal paragraph}

% to be able to preserve linearity accross the stages and to enable \emph{Lint} to
% preserve our sanity regarding linearity and eventually inform with linearity new
% transformations.

% A value allocated to be passed to a linear function and then never again used could
% bypass garbage collection.

@@

Now, with the following example, my interpretation of the current rule says that
the following case expression consumes the scrutinee twice because the case
binder is used. However, we know for sure that in the branch where z is used, z
is equal to False, and using False doesn't violate linearity since we can
unrestrictedly create nullary data constructors. A situation similar to this one
below happens after the CSE transformation.
\begin{code}
    case <complex expression> $\leadsto U$ of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }
\end{code}

@@

Taking the following example we can see that we don't know how to say
that x is indeed linear in one case and unrestricted in the other, even though
it is according to its type. We'd need some sort of coercion to coerce through
the multiplicity to the new one we uncover when we pattern match on the GADT
evidence (...)
